### 2016-06-27 stitched.tex
ENH:  More response items.

White matter hyperintensities (WMHs) are foci of abnormal signal
intensity in white matter regions seen with magnetic resonance imaging
(MRI). WMHs are associated with normal aging and have shown prognostic
value in neurological conditions such as traumatic brain injury (TBI).
The impracticality of manually quantifying these lesions limits their
clinical utility and motivates the utilization of machine learning
techniques for automated segmentation workflows. Herein, we develop a
concatenated random forest framework with image features for segmenting
WMHs in a TBI cohort. The framework is provided publicly through the
Advanced Normalization Tools (ANTs) and ANTsR toolkits. MR (3D FLAIR,
T2-, and T1-weighted) images from 24 service members and veterans
scanned in the Chronic Effects of Neurotrauma Consortium's (CENC)
observational study were acquired. Manual annotations were employed for
both training and evaluation using a leave-one-out strategy. Lesion load
and overlap evaluative comparisons are complimented by feature rankings
which showcase the utility of the concatenated approach. Our findings
suggest supervised learning methods may be applied to quantify WMHs on
routine brain imaging. Paired with correlative outcome data, supervised
learning methods may allow for identification of imaging features
predictive of diagnosis and prognosis in individual TBI patients.

@@

Random forests are conceptually straightforward\textsuperscript{9}. They
consist of ensembles of decision trees that are built from training
data. Once \textcolor{blue}{the ensemble of decision trees is}
constructed, data to be classified is ``pushed'' through each decision
tree resulting in a single classification ``vote'' per tree. These votes
are then used for regression or classification of the data. Although
decision trees had been extensively studied, the success of employing
collections of such weak learners for boosting machine learning
performance (e.g., AdaBoost\textsuperscript{18,19}) influenced the
similarly sytled conglomeration of decision trees into ``forests'' with
randomized node optimization\textsuperscript{20,21}. Finally,
Breiman\textsuperscript{9} improved accuracy by random sampling of
training data (i.e., ``bagging'') resulting in the current random forest
technique applied here.

@@

\textcolor{blue}{Figure X provides a graphical overview of the proposed workflow.  The major components
include offline generation of symmetric multimodal templates, the creation of
feature images from the training data which are then used to model the statistical
prediction using a concatenated random forest framework.  Once these offline steps
are performed, a new, unsegmented subject can then be processed using the proposed pipeline.}

@@

\textcolor{blue}{
Crucial to these supervised segmentation approaches are the creation and selection of
``features'' as input (i.e., feature images constructed from the training data)
in conjunction with expertly identified structures of interest
(i.e., WMHs) for model construction.} For the targeted application in
this work, tissue classification is performed at the voxelwise level. In
other words, each voxel within the region of interest is sent through
the ensemble of decision trees and receives a set of classification
votes from each tree thus permitting a regression or classification
solution. Since this procedure is performed at the voxelwise level,
intensity information alone is insufficient for good segmentation
performance due to the lack of spatial context. For example, as pointed
out in\textsuperscript{24}, higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated
lesion detection algorithms. Other potential confounds include MR signal
inhomogeneity and noise. Therefore, even though machine learning and
pattern recognition techniques are extremely powerful and have
significant potential, just as crucial to outcome is the creative
construction and deployment of salient feature images which we detail
below.

Supervised methodologies are uniquely characterized, in part, by the
feature images that are used to identify the regions of interest. In
Table 1, we provide a list and basic categorization of the feature
images used for the initial (i.e., Stage 1---more on the use of multiple
random forest stages below) segmentation of the WMHs. In addition Figure
1 provides a representation of a set of feature images for a single
subject analyzed in this work. Note that in this work we categorize the
brain parenchyma with seven labels:

@@

As mentioned previously, input for each subject comprises FLAIR, T1-,
and T2-weighted acquisitions. The T1 and T2 images are rigidly
registered to the FLAIR image using the open-source Advanced
Normalization Tools (ANTs)\textsuperscript{22}. The aligned images are
then preprocessed using the denoising algorithm of\textsuperscript{25}
followed by N4 bias correction\textsuperscript{26} which are then
normalized to the intensity range \([0,1]\). Although we could have used
an alternative intensity standardization algorithm
(e.g.,\textsuperscript{27}), we found that a simple linear rescaling
produced better results similar to previous work\textsuperscript{17}.

The T1 image is then processed via the ANTs brain extraction and normal
tissue segmentation pipelines\textsuperscript{28}. The result is a mask
delineating the brain parenchyma and probabilistic estimates of the CSF,
gray matter, white matter, deep gray matter, brain stem, and
cerebellum\textsuperscript{29}. These provide the expertly annotated
labels for the first six tissue labels given above. Tissue prior
probability maps for segmentation are from multi-model optimal symmetric
shape/intensity templates\textsuperscript{17} created from the public
MMRR data set\textsuperscript{30} (cf Figure 3).

Feature values include the preprocessed FLAIR, T1, and T2 image voxel
intensities. We also calculate a set of neighborhood statistics (mean,
standard deviation, and skewness) feature images using a Manhattan
radius of one voxel given the typical size of individual WMHs. For each
of the preprocessed images, we calculate the difference in intensities
with the corresponding warped template component. Previous success in
the international brain tumor segmentation
competition\textsuperscript{31} was based on an important set of
intensity features that were created from multi-modal templates
mentioned previously\textsuperscript{17} and listed in Table 1. We
employ the same strategy here.

@@

The segmentation probability images described above are used as feature
images to provide a spatial context for the random forest model
prediction step. Additional spatial contextual feature images include
the distance maps\textsuperscript{32} based on the csf, gray matter, and
deep gray matter images. These latter images are intended to help
distinguish white matter hyperintensities from false positives induced
by the partial voluming at the gray/white matter interface. A third set
of images are based on the voxel location within the space of the
template.
\textcolor{blue}{Similar feature images were used in}\textsuperscript{33}
\textcolor{blue}{although, unlike the proposed framework, this previous work lacks
normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.  To generate these images, the T1 image of each subject is
registered to the T1 template component using a B-spline variant}\textsuperscript{34}
\textcolor{blue}{of the well-known ANTs Symmetric Normalization (SyN) algorithm}\textsuperscript{35}.
\textcolor{blue}{Using the derived transforms, the template coordinate images are warped back to the space of the individual subject.}

@@

In previous brain tumor segmentation work\textsuperscript{17}, it was
demonstrated that a concatenated supervised approach, whereby the
prediction output from the first random forest model serves as partial
input for a second random forest model, can significantly improve
segmentation performance. We do the same thing for the work described
here where we employ two stacked random forests (or two ``stages''). The
Stage 1 feature images of the training data (as described previously)
are used to construct the Stage 1 model. The training data Stage 1
features are then used to produce the voxelwise ``voting maps'' (i.e.,
the classification count of each decision tree for each tissue label)
via the Stage 1 random forest model. All the Stage 1 features plus the
Stage 1 voting maps are used as input to the Stage 2 model. In addition,
we use the Stage 1 voting maps as tissue priors (i.e., probabilistic
estimates of the tissue spatial locations) for a second application of
the \(6\)-tissue segmentation algorithm with an additional Markov Random
Field spatial prior (MAP-MRF)\textsuperscript{29}.
\textcolor{blue}{In order to maximize the spatial information for the $n$-tissue segmentation process following the voxelwise RF classification of Stage 1, we use
all three aligned preprocessed images for multivariate segmentation during the
second stage.} The resulting seven posterior probability images
constitute a third additional feature image set for Stage 2.

@@

As pointed out in a recent comprehensive lesion segmentation
review\textsuperscript{36}, although the number of algorithms reported
in the literature is quite extensive, there were only four publicly
available segmentation algorithms at the time of writing this article.
In contrast to the current work, none are based on supervised learning.
As we did for our brain tumor segmentation
algorithm\textsuperscript{17}, all of the code described in this work is
publicly available through the open-source ANTs/ANTsR toolkits. Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R
statistical project) we use the \emph{randomForest}
package\textsuperscript{37} using the default settings with 2000 trees
per model and 500 randomly selected samples per label per image. Note
that we saw little variation in performance when these parameters were
changed (i.e.~up to 1000 random samples and as little as 1000 trees)
which is consistent with our previous experience.

@@

The resulting rankings for both Stages are given in Figures 4 and 5
where the values for the separate stages are averaged over the entire
corresponding model set. In addition, we track the variance for each
feature over all models to illustrate the stability of the chosen
features during the evaluation. This latter information is illustrated
as horizontal errors bars providing the \(95^{th}\) percentile Note that
the reader can cross reference Table 1 for identifying corresponding
feature types and names.

Additionally, it is interesting to note some of the other top performing
features for Stage 1. The contralateral difference FLAIR image is highly
discriminative over the set of evaluation random forest models (see
Figure 6). This accords with the known clinical relevance of FLAIR
images for identifying white matter hyperintensities and the fact that
such pathology does not typically manifest symmetrically in both
hemispheres. Interestingly, the posterior maps for the deep gray matter
are extremely important for accurate white matter hyperintensity
segmentation. Perhaps the spatial specification of deep gray matter aids
in the removal of false positives. Inspection of the bottom of the plots
demonstrates the lack of discriminating features associated with the T1
image which is also well-known in the clinical literature.

As described earlier, for Stage 2, we used the output random forest
voting maps from Stage 1 as both features themselves and as priors for
input to a Bayesian-based segmentation with an additional MRF spatial
prior. In Figure 5, the voting maps are labeled as
``\emph{RFStage1VotingMaps}'' where the final numeral is associated with
the brain parenchymal labeling given previously. Similarly, the
additional RF prior segmentation feature probability maps are labeled as
``\emph{RFBrainSegmentationPosteriors}''. The Stage 2 feature importance
plot follows similar trends as that for Stage 1 with the T1 images not
contributing much to the identification of white matter hyperintensity
voxels. The initial voting maps from Stage 1 are extremely important
with the top 3 being the estimated locations of the 1) gray matter, 2)
white matter, and 3) white matter hyperintensities. Since these tissue
type can be conflated based on intensity alone it is intuitive that such
features would be important.

@@

In Figure 7 are the segmentation comparisons derived from manual
segmentations of the same data. Despite the large variability
characteristic with manual labelings in related
fields\textsuperscript{36,38,39}, such labelings are characteristic of
current clinical practices and the methodology proposed herein is
readily adapted to refinements in training data.

On the left of Figure 7 are the improvement in Dice
values\textsuperscript{40}, over all white matter hyperintensities when
comparing the segmentations between the two stages where the sum is
taken over all individually labeled manual, \(T_r\), and automated,
\(S_r\), lesions and \(\cap\) represents the intersection between the
manual/automated lesion pair. Performing the second round of supervised
learning improves these Dice values. One can also note from the right
side of Figure 6 that the total lesion load volume illustrates a few
subjects that are severe outliers in terms of the number of false
positives. The second round helps to correct this issue.

@@

Two major meta-analyses of WMHs have been published covering the periods
prior to 2010\textsuperscript{41} and after\textsuperscript{42}. The
earlier meta-analysis covered 53 longitudinal studies that included
samples of high-risk populations, i.e., patients selected for a specific
disease or condition such as hypertension, whereas other studies
recruited samples of the general population. Longitudinal studies of
samples representative of the general population are more relevant to
the focus of the present paper. Debette \& Markus\textsuperscript{41}
found that the presence of WMHs was related to subsequent cognitive
decline, a higher risk of developing dementia, stroke, and of mortality.
Lesion volume at baseline was also predictive of cognitive decline.
Limitations of this meta-analysis include heterogeneity in the method of
measuring WMHs; some studies used automated volumetric measurement,
whereas others used a visual rating scale. The studies analyzed by
Debette \& Markus were limited to the occurrence of one of the
aforementioned conditions which they analyzed by hazard ratios.

The more recent meta-analysis by Kloppenborg et al.\textsuperscript{42}
of 23 cross-sectional studies reporting MRI and concurrent
neuropsychological results in patients with heterogeneous diagnoses but
without previously diagnosed cognitive impairment, found that WMHs were
associated with cognitive deficit (effect size of -0.10, 95\% CI: -0.13
to -0.08) after controlling for age. These studies also differed in the
metric used to measure the WMHs, including volume, \% of total
intracranial volume, and a visual rating score. The effect size for the
association with cognitive deficit in these cross-sectional studies did
not differ significantly across various cognitive domains or the method
of measuring lesion volume. Among eight longitudinal studies analyzed by
Kloppenborg et al that included a follow-up MRI and also controlled for
age, the effect size for the association of progression in WMHs and
cognitive impairment was -0.16 (95\% CI:-0.27 to -0.09). This
association was stronger for attention and executive function than for
memory and processing speed. Although baseline WMHs were predictive of
cognitive deficit at follow-up in the seven studies which did not repeat
MRI, the effect size was smaller {[}-0.10 (95\% CI: 0.13 to -0.05) than
in the longitudinal studies that calculated progression in WMHs. In
summary, progression of WMHs seen on repeat MRI has a stronger relation
to cognitive deficit than concurrent imaging findings. These
meta-analyses support the rationale for repeating an MRI in patients
younger than 50 years whose initial scan shows WMHs.

@@

The currently described tool is just one example of how ``supervised
learning'' algorithms might be applied to aid in the diagnosis of TBI
and other disease processes through the specific identification of
features predictive of a given disease state. It is an important
demonstration of the potential power of these analytical approaches in
the rapid but comprehensive mining of information from neuroimaging
examinations. Supervised learning algorithms are presently employed
across a wide variety of settings for the rapid identification of
predictive imaging features\textsuperscript{43--46}. Automobile
manufacturers utilize these types of approaches to equip self-driving
vehicles to recognize and respond to unique external surroundings
through the identification of visual information sufficiently similar to
previously assimilated training data\textsuperscript{47,48}. Similarly,
in the context of the neuroimaging assessments, deep learning approaches
may allow for the rapid identification of information predictive of
disease state in an individual patient. These approaches have been
applied to the segmentation of macroscopically visible
structures\textsuperscript{43--46}. Additionally, these approaches might
be applied to the interrogation of imaging data in the individual
patient with a primary quantitative output metrics to include sequences
such as diffusion tensor imaging (DTI) and its variants, functional
connectivity, perfusion weighted imaging, and cortical thickness
assessments. At present, these advanced neuroimaging sequences are
confined to cohort-based research studies due to the lack of available
analytical tools to assess the information in the setting of the
individual patient\textsuperscript{49}. Application of deep learning
approaches in the context of data with primary quantitative outputs will
require large scale normative and disease specific databases. Building
these large scale imaging libraries is resource intensive and requires a
multi-center approach with harmonized scanners between sites and
correlative non-imaging clinical data. Large scale TBI data is becoming
increasingly available through activities such as the Chronic Effects of
Neurotrauma Consortium (CENC), Transforming Research and Clinical
Knowledge in TBI (TRACK-TBI), Collaborative European Neurotrauma
Effectiveness Research in TBI (CENTER-TBI), Department of Defense
Alzheimer's Disease Neuroimaging Initiative (DOD-ADNI), and other data
being consolidated through FITBIR. In concert with any available high
quality normative neuroimaging data, deep learning algorithms may be
well positioned to help transform how neuroimaging is interpreted for
the clinical management of patients with this disease process.

@@

\textbf{Figure X:}
\textcolor{blue}{Workflow illustration for the proposed pipeline.  Processing of the multi-modal
input MRI for a single subject, using the multi-modal symmetric template, results in
the generation of the feature images.  These feature images are used as input to the
Stage 1 RF model producing the initial RF probability map estimates.  The Stage 1
voting maps, the original feature images, and the Stage 2 RF model result in the
final voting maps which includes the WMH probability estimate.  Note that the RF models
are constructed once from a set of training data which are processed using the
same feature-construction pipeline as the single-subject input MRI.}

\textbf{Figure 1:} Representation of Stage 1 feature images for subject
01C1019. The FLAIR, T1-, and T2-weighted images are rigidly
pre-aligned\textsuperscript{22} to the space of the T1 image. The three
modality images are then preprocessed (N4 bias
correction\textsuperscript{26} and adaptive
denoising\textsuperscript{25}) followed by application of standard ANTs
brain extraction and \(n\)-tissue segmentation protocols using the MMRR
symmetric template and corresponding priors\textsuperscript{28} applied
to the T1 image. The feature images are then generated for voxelwise
input to the RF model which results in the voting maps illustrated on
the right. This gives a probabilistic classification of tissue type. Not
shown are the probability and voting images for the brain stem and
cerebellum.

\textbf{Figure 2:} Sample FLAIR acquisition image slices showing both
manual and random forest segmentations for both stages obtained during
the leave-one-out evaluation. Manual segmentations were performed by one
of the authors and provided the ground truth WMH labels for training the
random forest models.

\textbf{Figure 3:} Canonical views of the mutlivariate, bilaterally
symmetric template constructed from the MMRR data
set\textsuperscript{30} (only shown are the FLAIR, T1, and T2
modalities--- the components relevant for this work). Template
construction is detailed in\textsuperscript{17}. These images are
important for specific intensity-based features.

>>>>

White matter hyperintensities (WMHs) are foci of abnormal signal
intensity in white matter regions seen with magnetic resonance imaging
(MRI). WMHs are associated with normal aging and have shown prognostic
value in neurological conditions such as traumatic brain injury (TBI).
The impracticality of manually quantifying these lesions limits their
clinical utility and motivates the utilization of machine learning
techniques for automated segmentation workflows. Herein, we develop a
concatenated random forest framework with image features for segmenting
WMHs in a TBI cohort. The framework is built upon the Advanced
Normalization Tools (ANTs) and ANTsR toolkits. MR (3D FLAIR, T2-, and
T1-weighted) images from 24 service members and veterans scanned in the
Chronic Effects of Neurotrauma Consortium's (CENC) observational study
were acquired. Manual annotations were employed for both training and
evaluation using a leave-one-out strategy. Lesion load and overlap
evaluative comparisons are complimented by feature rankings which
showcase the utility of the concatenated approach. Our findings suggest
supervised learning methods may be applied to quantify WMHs on routine
brain imaging. Paired with correlative outcome data, supervised learning
methods may allow for identification of imaging features predictive of
diagnosis and prognosis in individual TBI patients.

@@

Random forests are conceptually straightforward\textsuperscript{9}.
\textcolor{blue}{The basic component of
the random forest paradigm is the ``decision tree'' often represented by a flowchart or
graph where internal nodes represent ``tests'', or decisions, and the edges represent the outcome of those tests.  The final, or end, nodes represent the various classsifications produced by traversal through the decision tree.  For the proposed application, individual
voxels (and their corresponding feature values) are introduced at the root of a particular
decision tree and traverse the edges and internal nodes ultimately ending up at one of
the classification nodes according to the tests at each internal node.  A single random
forest model will consist of many such trees (often refered to as an ``ensemble'').}
\textcolor{blue}{Although decision trees had been extensively studied, the success of employing collections of such weak learners for boosting machine learning performance
(e.g., AdaBoost)}\textsuperscript{18,19}
\textcolor{blue}{influenced the similarly sytled conglomeration of decision trees into ``forests'' with randomized node optimization}\textsuperscript{20,21}.
\textcolor{blue}{Finally, Breiman}\textsuperscript{9}
\textcolor{blue}{improved accuracy by random sampling of training data (i.e., ``bagging'') resulting
in the current random forest technique applied here.}
\textcolor{blue}{As voxels and their feature values are ``pushed'' through each decision tree in the forest,  votes for each label are accumulated and converted to probability values for all classification possiblilities at
each voxel location.}

@@

\textcolor{blue}{Figure 1 provides a graphical overview of the proposed workflow.  The major components
include offline generation of symmetric multimodal templates, the creation of
feature images from the training data which are then employed for statistical
prediction using a concatenated random forest framework.  This framework involves
the use of two random forest models where the outcome (i.e., the tissue
probability estimates) of the first RF model application,
which we denote as ``Stage1 '', is used as input (along with the original set of
feature images) to a refinement RF model segmentation which we denote as ``Stage 2''.
Once these offline steps
are performed, a new, unsegmented subject can then be processed using the proposed pipeline.}

\subsubsection{Symmetric multi-modal
templates}\label{symmetric-multi-modal-templates}

\textcolor{blue}{Following}\textsuperscript{24}
\textcolor{blue}{and}\textsuperscript{17},
\textcolor{blue}{optimally derived templates
serve for both brain tissue segmentation (for the derivation of tissue prior probability maps) and generation of asymmetry feature images.
For this work we use the multi-modal data available from the public MMRR data set}\textsuperscript{25}\textcolor{blue}{.  We used all 46 multi-modal acquisitions of that study to produce a multi-modal template according to the procedure described in}\textsuperscript{26}
\textcolor{blue}{which
results in a mean (in terms of both shape and intensity) multivariate template representing the entire cohort.  Mid-canonical slices of the FLAIR, T1, and T2 components are illustrated in Figure 2.}

@@

\textcolor{blue}{
Crucial to these supervised segmentation approaches are the creation and selection of
``features'' as input (i.e., feature images constructed from the training data)
in conjunction with expertly identified structures of interest
(i.e., WMHs) for model construction.} For the targeted application in
this work, tissue classification is performed at the voxelwise level. In
other words, each voxel within the region of interest is sent through
the ensemble of decision trees and receives a set of classification
votes from each tree thus permitting a regression or classification
solution. Since this procedure is performed at the voxelwise level,
intensity information alone is insufficient for good segmentation
performance due to the lack of spatial context. For example, as pointed
out in\textsuperscript{27}, higher intensities can be found at the
periventricular caps in normal subjects which often confounds automated
lesion detection algorithms. Other potential confounds include MR signal
inhomogeneity and noise. Therefore, even though machine learning and
pattern recognition techniques are extremely powerful and have
significant potential, just as crucial to outcome is the creative
construction and deployment of salient feature images which we detail
below.

Supervised methodologies are uniquely characterized, in part, by the
feature images that are used to identify the regions of interest. In
Table 1, we provide a list and basic categorization of the feature
images used for the initial (i.e., Stage 1---more on the use of multiple
random forest stages below) segmentation of the WMHs. In addition Figure
3 provides a representation of a set of feature images for a single
subject analyzed in this work. Note that in this work we categorize the
brain parenchyma with seven labels:

@@

As mentioned previously, input for each subject comprises FLAIR, T1-,
and T2-weighted acquisitions. The T1 and T2 images are rigidly
registered to the FLAIR image using the open-source Advanced
Normalization Tools (ANTs)\textsuperscript{22}. The aligned images are
then preprocessed using the denoising algorithm of\textsuperscript{28}
followed by N4 bias correction\textsuperscript{29} which are then
normalized to the intensity range \([0,1]\). Although we could have used
an alternative intensity standardization algorithm
(e.g.,\textsuperscript{30}), we found that a simple linear rescaling
produced better results similar to previous work\textsuperscript{17}.

The T1 image is then processed via the ANTs brain extraction and normal
tissue segmentation pipelines\textsuperscript{24}. The result is a mask
delineating the brain parenchyma and probabilistic estimates of the CSF,
gray matter, white matter, deep gray matter, brain stem, and
cerebellum\textsuperscript{31}. These provide the expertly annotated
labels for the first six tissue labels given above. Tissue prior
probability maps for segmentation are from multi-model optimal symmetric
shape/intensity templates\textsuperscript{17} created from the public
MMRR data set\textsuperscript{25} (cf Figure 2).

Feature values include the preprocessed FLAIR, T1, and T2 image voxel
intensities. We also calculate a set of neighborhood statistics (mean,
standard deviation, and skewness) feature images using a Manhattan
radius of one voxel given the typical size of individual WMHs. For each
of the preprocessed images, we calculate the difference in intensities
with the corresponding warped template component. Previous success in
the international brain tumor segmentation
competition\textsuperscript{32} was based on an important set of
intensity features that were created from multi-modal templates
mentioned previously\textsuperscript{17} and listed in Table 1. We
employ the same strategy here.

@@

The segmentation probability images described above are used as feature
images to provide a spatial context for the random forest model
prediction step. Additional spatial contextual feature images include
the distance maps\textsuperscript{33} based on the csf, gray matter, and
deep gray matter images. These latter images are intended to help
distinguish white matter hyperintensities from false positives induced
by the partial voluming at the gray/white matter interface. A third set
of images are based on the voxel location within the space of the
template.
\textcolor{blue}{Similar feature images were used in}\textsuperscript{34}
\textcolor{blue}{although, unlike the proposed framework, this previous work lacks
normalization to the standard
coordinate system provided by the template to dramatically improve spatial specificity
across all subjects.  To generate these images, the T1 image of each subject is
registered to the T1 template component using a B-spline variant}\textsuperscript{35}
\textcolor{blue}{of the well-known ANTs Symmetric Normalization (SyN) algorithm}\textsuperscript{36}.
\textcolor{blue}{Using the derived transforms, the template coordinate images are warped back to the space of the individual subject.}

@@

In previous brain tumor segmentation work\textsuperscript{17}, it was
demonstrated that a concatenated supervised approach, whereby the
prediction output from the first random forest model serves as partial
input for a second random forest model, can significantly improve
segmentation performance. We do the same thing for the work described
here where we employ two stacked random forests (or two ``stages''). The
Stage 1 feature images of the training data (as described previously)
are used to construct the Stage 1 model. The training data Stage 1
features are then used to produce the voxelwise ``voting maps'' (i.e.,
the classification count of each decision tree for each tissue label)
via the Stage 1 random forest model. All the Stage 1 features plus the
Stage 1 voting maps are used as input to the Stage 2 model. In addition,
we use the Stage 1 voting maps as tissue priors (i.e., probabilistic
estimates of the tissue spatial locations) for a second application of
the \(6\)-tissue segmentation algorithm with an additional Markov Random
Field spatial prior (MAP-MRF)\textsuperscript{31}.
\textcolor{blue}{In order to maximize the spatial information for the $n$-tissue segmentation process following the voxelwise RF classification of Stage 1, we use
all three aligned preprocessed images for multivariate segmentation during the
second stage.} The resulting seven posterior probability images
constitute a third additional feature image set for Stage 2.

@@

As pointed out in a recent comprehensive lesion segmentation
review\textsuperscript{37}, although the number of algorithms reported
in the literature is quite extensive, there were only four publicly
available segmentation algorithms at the time of writing this article.
In contrast to the current work, none are based on supervised learning.
As we did for our brain tumor segmentation
algorithm\textsuperscript{17}, all of the code described in this work is
publicly available through the open-source ANTs/ANTsR toolkits. Through
ANTsR (an add-on toolkit which, in part, bridges ANTs and the R
statistical project) we use the \emph{randomForest}
package\textsuperscript{38} using the default settings with 2000 trees
per model and 500 randomly selected samples per label per image. Note
that we saw little variation in performance when these parameters were
changed (i.e.~up to 1000 random samples and as little as 1000 trees)
which is consistent with our previous experience.

@@

The resulting rankings for both Stages are given in Figures 5 and 6
where the values for the separate stages are averaged over the entire
corresponding model set. In addition, we track the variance for each
feature over all models to illustrate the stability of the chosen
features during the evaluation. This latter information is illustrated
as horizontal errors bars providing the \(95^{th}\) percentile Note that
the reader can cross reference Table 1 for identifying corresponding
feature types and names.

Additionally, it is interesting to note some of the other top performing
features for Stage 1. The contralateral difference FLAIR image is highly
discriminative over the set of evaluation random forest models (see
Figure 7). This accords with the known clinical relevance of FLAIR
images for identifying white matter hyperintensities and the fact that
such pathology does not typically manifest symmetrically in both
hemispheres. Interestingly, the posterior maps for the deep gray matter
are extremely important for accurate white matter hyperintensity
segmentation. Perhaps the spatial specification of deep gray matter aids
in the removal of false positives. Inspection of the bottom of the plots
demonstrates the lack of discriminating features associated with the T1
image which is also well-known in the clinical literature.

As described earlier, for Stage 2, we used the output random forest
voting maps from Stage 1 as both features themselves and as priors for
input to a Bayesian-based segmentation with an additional MRF spatial
prior. In Figure 6, the voting maps are labeled as
``\emph{RFStage1VotingMaps}'' where the final numeral is associated with
the brain parenchymal labeling given previously. Similarly, the
additional RF prior segmentation feature probability maps are labeled as
``\emph{RFBrainSegmentationPosteriors}''. The Stage 2 feature importance
plot follows similar trends as that for Stage 1 with the T1 images not
contributing much to the identification of white matter hyperintensity
voxels. The initial voting maps from Stage 1 are extremely important
with the top 3 being the estimated locations of the 1) gray matter, 2)
white matter, and 3) white matter hyperintensities. Since these tissue
type can be conflated based on intensity alone it is intuitive that such
features would be important.

@@

In Figure 8 are the segmentation comparisons derived from manual
segmentations of the same data. Despite the large variability
characteristic with manual labelings in related
fields\textsuperscript{37,39,40}, such labelings are characteristic of
current clinical practices and the methodology proposed herein is
readily adapted to refinements in training data.

On the left of Figure 8 are the improvement in Dice
values\textsuperscript{41}, over all white matter hyperintensities when
comparing the segmentations between the two stages where the sum is
taken over all individually labeled manual, \(T_r\), and automated,
\(S_r\), lesions and \(\cap\) represents the intersection between the
manual/automated lesion pair. Performing the second round of supervised
learning improves these Dice values. One can also note from the right
side of Figure 7 that the total lesion load volume illustrates a few
subjects that are severe outliers in terms of the number of false
positives. The second round helps to correct this issue.

@@

Two major meta-analyses of WMHs have been published covering the periods
prior to 2010\textsuperscript{42} and after\textsuperscript{43}. The
earlier meta-analysis covered 53 longitudinal studies that included
samples of high-risk populations, i.e., patients selected for a specific
disease or condition such as hypertension, whereas other studies
recruited samples of the general population. Longitudinal studies of
samples representative of the general population are more relevant to
the focus of the present paper. Debette \& Markus\textsuperscript{42}
found that the presence of WMHs was related to subsequent cognitive
decline, a higher risk of developing dementia, stroke, and of mortality.
Lesion volume at baseline was also predictive of cognitive decline.
Limitations of this meta-analysis include heterogeneity in the method of
measuring WMHs; some studies used automated volumetric measurement,
whereas others used a visual rating scale. The studies analyzed by
Debette \& Markus were limited to the occurrence of one of the
aforementioned conditions which they analyzed by hazard ratios.

The more recent meta-analysis by Kloppenborg et al.\textsuperscript{43}
of 23 cross-sectional studies reporting MRI and concurrent
neuropsychological results in patients with heterogeneous diagnoses but
without previously diagnosed cognitive impairment, found that WMHs were
associated with cognitive deficit (effect size of -0.10, 95\% CI: -0.13
to -0.08) after controlling for age. These studies also differed in the
metric used to measure the WMHs, including volume, \% of total
intracranial volume, and a visual rating score. The effect size for the
association with cognitive deficit in these cross-sectional studies did
not differ significantly across various cognitive domains or the method
of measuring lesion volume. Among eight longitudinal studies analyzed by
Kloppenborg et al that included a follow-up MRI and also controlled for
age, the effect size for the association of progression in WMHs and
cognitive impairment was -0.16 (95\% CI:-0.27 to -0.09). This
association was stronger for attention and executive function than for
memory and processing speed. Although baseline WMHs were predictive of
cognitive deficit at follow-up in the seven studies which did not repeat
MRI, the effect size was smaller {[}-0.10 (95\% CI: 0.13 to -0.05) than
in the longitudinal studies that calculated progression in WMHs. In
summary, progression of WMHs seen on repeat MRI has a stronger relation
to cognitive deficit than concurrent imaging findings. These
meta-analyses support the rationale for repeating an MRI in patients
younger than 50 years whose initial scan shows WMHs.

@@

The currently described tool is just one example of how ``supervised
learning'' algorithms might be applied to aid in the diagnosis of TBI
and other disease processes through the specific identification of
features predictive of a given disease state. It is an important
demonstration of the potential power of these analytical approaches in
the rapid but comprehensive mining of information from neuroimaging
examinations. Supervised learning algorithms are presently employed
across a wide variety of settings for the rapid identification of
predictive imaging features\textsuperscript{44--47}. Automobile
manufacturers utilize these types of approaches to equip self-driving
vehicles to recognize and respond to unique external surroundings
through the identification of visual information sufficiently similar to
previously assimilated training data\textsuperscript{48,49}. Similarly,
in the context of the neuroimaging assessments, deep learning approaches
may allow for the rapid identification of information predictive of
disease state in an individual patient. These approaches have been
applied to the segmentation of macroscopically visible
structures\textsuperscript{44--47}. Additionally, these approaches might
be applied to the interrogation of imaging data in the individual
patient with a primary quantitative output metrics to include sequences
such as diffusion tensor imaging (DTI) and its variants, functional
connectivity, perfusion weighted imaging, and cortical thickness
assessments. At present, these advanced neuroimaging sequences are
confined to cohort-based research studies due to the lack of available
analytical tools to assess the information in the setting of the
individual patient\textsuperscript{50}. Application of deep learning
approaches in the context of data with primary quantitative outputs will
require large scale normative and disease specific databases. Building
these large scale imaging libraries is resource intensive and requires a
multi-center approach with harmonized scanners between sites and
correlative non-imaging clinical data. Large scale TBI data is becoming
increasingly available through activities such as the Chronic Effects of
Neurotrauma Consortium (CENC), Transforming Research and Clinical
Knowledge in TBI (TRACK-TBI), Collaborative European Neurotrauma
Effectiveness Research in TBI (CENTER-TBI), Department of Defense
Alzheimer's Disease Neuroimaging Initiative (DOD-ADNI), and other data
being consolidated through FITBIR. In concert with any available high
quality normative neuroimaging data, deep learning algorithms may be
well positioned to help transform how neuroimaging is interpreted for
the clinical management of patients with this disease process.

@@

\textbf{Figure 1:}
\textcolor{blue}{Workflow illustration for the proposed pipeline.  Processing of the multi-modal
input MRI for a single subject, using the multi-modal symmetric template, results in
the generation of the feature images.  These feature images are used as input to the
Stage 1 RF model producing the initial RF probability map estimates.  The Stage 1
voting maps, the original feature images, and the Stage 2 RF model result in the
final voting maps which includes the WMH probability estimate.  Note that the RF models
are constructed once from a set of training data which are processed using the
same feature-construction pipeline as the single-subject input MRI.}

\textbf{Figure 2:} Canonical views of the mutlivariate, bilaterally
symmetric template constructed from the MMRR data
set\textsuperscript{25} (only shown are the FLAIR, T1, and T2
modalities--- the components relevant for this work). Template
construction is detailed in\textsuperscript{17}. These images are
important for asymmetry-based features.

\textbf{Figure 3:} Representation of Stage 1 feature images for subject
01C1019. The FLAIR, T1-, and T2-weighted images are rigidly
pre-aligned\textsuperscript{22} to the space of the T1 image. The three
modality images are then preprocessed (N4 bias
correction\textsuperscript{29} and adaptive
denoising\textsuperscript{28}) followed by application of standard ANTs
brain extraction and \(n\)-tissue segmentation protocols using the MMRR
symmetric template and corresponding priors\textsuperscript{24} applied
to the T1 image. The feature images are then generated for voxelwise
input to the RF model which results in the voting maps illustrated on
the right. This gives a probabilistic classification of tissue type. Not
shown are the probability and voting images for the brain stem and
cerebellum.

\textbf{Figure 4:} Sample FLAIR acquisition image slices showing both
manual and random forest segmentations for both stages obtained during
the leave-one-out evaluation. Manual segmentations were performed by one
of the authors and provided the ground truth WMH labels for training the
random forest models.

\textbf{Figure 5:} Average \emph{MeanDecreaseAccuracy} plots generated
from the creation of all 24 random forest models for Stage 1 during the
leave-one-out evaluation. These plots are useful in providing a
quantitative assessment of the predictive importance of each feature.
Features are ranked in descending order of importance. The horizontal
error bars provide the \(95^{th}\) percentile and illustrate the
stability of the feature importance across the leave-one-out models. At
this initial stage only 31 feature images are used.
